CrypTool Script \\ Mathematics and Cryptography

### 2002-06-03 script/en/chapters/cryptomethods.tex
New script directory and file layout (only English for now)

\documentclass[a4paper,11pt]{article}
% Da man keine Unterst"utzung der deutschen Sprache in diesem Fall braucht!!
% \usepackage[german]{babel}

\setlength{\fboxrule}{.5mm}
\setlength{\fboxsep}{1.75mm}
\setlength{\footnotesep}{6pt}
\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amsthm}
\usepackage{makeidx}
\usepackage[pdftex]{graphicx}
% \usepackage{graphicx}
\makeindex
\hypersetup{colorlinks=true}
\usepackage[flushmargin]{footmisc}

\setlength{\textwidth}{460 pt}
\hoffset-0.7in
\newtheorem{definition}{Definition}
%\newcommand{\href}[2]{#2}
\newcommand{\hypertag}[2]{#1}
%\newcommand{\hyperlink}[2]{#2}
%\newcommand{\hypertarget}[2]{#2}
\newtheorem{theorem}{Theorem}
\newenvironment{Proof}[1]
  {\textbf{Proof #1} \\}
  {\hfill$\Box$ \\}

%\pdfinfo{ /Title (CrypTool script)
%/Creator (TeX)
%/Producer (pdfTeX 0.14a)
%/Author (Deutsch Bank AG)
%/CreationDate (D:20000920201000)
%}

\title{ CrypTool Script \\ Mathematics and Cryptography }

\author{
(c) The authors, 1998-2002 \\
Frankfurt am Main
}

\begin{document}
\pagestyle{plain}
\setlength{\fboxrule}{.5mm}
\setlength{\fboxsep}{1.75mm}
\setlength{\footnotesep}{6pt}
\addtolength{\footskip}{8pt}
%\setlength{\footskip}{4cm}
%\renewcommand{\footnoterule}{\parindent0cm\rule{13cm}{.1pt}\vspace{.2cm}}
%Formatierung der Fu"snoten

%space between text and footnote 
\renewcommand\footnoterule{%
  \vspace{2em}%   <-- one line space between text and footnoterule
%\kern-3\p@
  \hrule width .4\columnwidth
 \vspace{4pt}
%kern 2.6\p@
}

%\long\def\@makefntext#1{%
%    \parindent 1em%
 %   \noindent
%    \hbox to 1.8em{\hss\@makefnmark}#1}

\maketitle

\parskip 4pt
\vskip + 30 pt
{
In this CrypTool script you will find predominantly mathematically oriented
information for using encryption procedures. The main chapters have been written
by {\em various authors} and are therefore independent from one another. At the end of
most chapters you will find literature and URLs.

You will receive information about the principles of symmetrical and
asymmetrical encryption. A large section of the script is dedicated to the
fascinating topic of prime numbers. Using numerous examples, the elementary number
theory and modular arithmetic are introduced and applied in an exemplary 
manner in the RSA procedure.  After this, one receives insight into the mathematical ideas
behind modern cryptography.

%You will also obtain an overview of the
%mathematical ideas behind modern cryptography.

A further chapter is devoted to digital signatures, which are an essential
component of e-business applications. The last chapter --- elliptic curves ---
fits in well with this. The mathematics of elliptic curves forms the basis for
rapid cryptographic algorithms for digital signatures; these algorithms are
extremely well suited for implementation on smartcards.

Whereas the program CrypTool teaches you how to use cryptography in practice, the
script provides those interested in the subject with a deeper understanding of
the mathematical algorithms used -- trying to do it didactically as good as possible.

The {\em authors} Bernhard Esslinger, Bartol Filipovic, Henrik Koy and Roger Oyono
would like to take this opportunity to thank their colleagues in the company and
at the universities of Frankfurt, Giessen and Karlsruhe. They particularly thank
Dr. Peer Wichmann from the Karlsruhe computer science research centre
(Forschungszentrum Informatik, FZI) for his down-to-earth support.

\enlargethispage{0.5cm}
As at CrypTool, the quality of the script is enhanced through your suggestions
and ideas for improvement. We look forward to your feedback.

\vskip +7 pt \noindent
You will find the current version of CrypTool under \newline
  \href{http://www.CrypTool.de}{\tt http://www.CrypTool.de},~~
  \href{http://www.cryptool.com}{\tt http://www.CrypTool.com}~~ or~~
  \href{http://www.cryptool.org}{\tt http://www.CrypTool.org}.
{
\vskip + 7 pt \noindent
The contact persons for this free tool are listed in the readme file for
CrypTool.}
}

\newpage
\tableofcontents
\addcontentsline{toc}{section}{Contents}
\newpage
\normalsize
\parindent0cm
\parskip4pt

\section*{Introduction}  \addcontentsline{toc}{section}{Introduction}

This script is delivered together with CrypTool.

CrypTool is a program with an extremely comprehensive online help enabling you
to use and analyse cryptographic procedures.\par \vskip + 3pt

For the conventional encryption procedures, automatic analyses are available in CrypTool
that enable you to determine the key with knowledge of the encrypted document
and, possibly, further information (non-encrypted document or language of the
document).\par \vskip + 3pt

CrypTool was developed during the End-User Awareness program in order to
increase employee awareness of IT security and provide them with a deeper
understanding of the term security.\par \vskip + 3pt

A further aim was to enable users to understand the cryptographic procedures
implemented in the Deutsche Bank. In this way, using CrypTool as a reliable
reference implementation of the various encryption procedures (because of using the industry-proven SECUDE Library),
you can test the encryption implemented in other programs. \par \vskip + 3pt

Because the articles in this script are largely self-contained, this script can also
be read independantly of CrypTool.

The {\em authors} tried to describe cryptography for a broad audience --
without being mathematically incorrect. We believe, that this didactical pretention
is the way to promote the awareness for IT security and the readiness to use standardised modern cryptography.

@@

One encryption procedure has been proved to be secure --- the {\em One-Time-Pad}.
\index{One­Time­Pad} However, this procedure has several practical
disadvantages (the key used must be selected randomly and must be just as long
as the message to be protected), which means that it is hardly used except in
closed environments such as for the hot wire between Moscow and Washington.\par \vskip + 3pt

@@

The most well-known symmetric encryption procedure is the \index{DES} DES--
algorithm. The DES-algorithm has been developed by IBM in collaboration with the
National Security Agency \index{NSA} (NSA), and was published as a standard in
1975. Despite the fact that the procedure is relatively old, no effective attack
on it has yet been detected. The most effective way of attacking consists of
testing all possible keys until the right one is found ({\em brute--force--attack}).
\index{Attack!brute-force} Due to the relatively short key length of
effectively 56 bits (64 bits, which however include 8 parity bits), numerous
messages encrypted using DES have in the past been broken. Therefore, the
procedure can now only be considered to be conditionally secure. Symmetric
alternatives to the DES procedure include the IDEA \index{IDEA} or Triple DES
algorithms.\par \vskip + 3pt

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% C H A P T E R   T W O
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newpage
\section{Prime numbers}\hypertarget{Kapitel_2}{}
(Bernhard Esslinger, besslinger@web.de, May 1999, Updates Nov. 2000, December 2001)

%\paragraph{What are prime numbers?}
\subsection*{What are prime numbers?}
\addcontentsline{toc}{subsection}{~~~~~~What are prime numbers?}
\index{Prime number} \index{Numbers!prime numbers}
Prime numbers are whole, positive numbers greater than or equal to $2$ that can
only be divided by 1 and themselves. All other natural numbers greater than or
equal to $2$ can be formed by multiplying prime numbers.

The {\em natural} \index{Numbers} numbers $\mathbb{N}=\{1, 2, 3, 4,\cdots \}$ thus comprise 
\begin{itemize}
   \item the number $1$ (the unit value)
   \item the primes and
   \item the composite numbers.
\end{itemize}

Prime numbers are particularly important for 3 reasons:
\begin{itemize}
  \item In number theory, they are considered to be the basic components of
natural numbers, upon which numerous brilliant mathematical ideas are based.
  \item They are of extreme practical importance in modern
\index{Cryptography!modern} cryptography (public key \index{Cryptography!public 
key} cryptography). The most common public key procedure, invented at the end of
the 1970's, is \index{RSA} RSA encryption. Only using (large) prime numbers for
particular parameters can you guarantee that an algorithm is secure, both for
the RSA procedure and for even more modern procedures (digital
\index{Signatures!digital} signature, elliptic curves).
  \item The search for the largest known prime numbers does not have any
practical usage known to date, but requires the best computers, is an excellent
benchmark (possibility for determining the performance of computers) and leads
to new calculation methods on many computers \\ (see also:
\href{http://www.mersenne.org/prime.htm}{\tt
http://www.mersenne.org/prime.htm}).
\end{itemize}
Many people have been fascinated by prime numbers over the past two millenniums.
Ambition to make new discoveries about prime numbers has often resulted in
brilliant ideas and conclusions. The following section provides an easily
comprehensible introduction to the basics of prime numbers. We will also explain
what is known about the distribution (density, number of prime numbers in
particular intervals) of prime numbers and how prime number tests work.

\subsection{Prime numbers in mathematics}

Every whole number has a factor. The number 1 only has one factor,
itself, whereas the number $12$ has the six factors $1, 2, 3, 4,
6, 12$. Many numbers can only be divided by themselves and by $1$.
With respect to multiplication, these are the ''atoms'' in the
area of numbers. Such numbers are called prime numbers.

In mathematics, a slightly different (but equivalent) definition is used.

\begin{definition}\label{def-pz-prime}
A whole number $p \in {\bf N}$ is called prime \index{Numbers!prime numbers} if
$p > 1$ and $p$ only possesses the trivial factors $\pm 1$ and $\pm p$.
\end{definition}

By definition, the number $1$ is not a prime number. In the following sections,
$p$ will always denote a prime number.

The sequence of prime numbers starts with $$ 2,~ 3,~ 5,~ 7, ~ 11, ~ 13, ~
17, ~ 19, ~ 23, ~ 29, ~ 31, ~ 37, ~ 41, ~ 43, ~ 47, ~ 53, ~ 59, ~ 61,
~ 67, ~ 71, ~ 73, ~ 79, ~ 83, ~ 89, ~ 97, \cdots . $$
The first 100 numbers include precisely 25 prime numbers. After this, the percentage of primes
constantly decreases. Prime numbers can be factorised in a unique {\em trivial}
way: $$5 = 1 \cdot 5,\quad  17 = 1 \cdot 17, \quad 1,013 = 1 \cdot 1,013,  \quad
1,296,409 = 1 \cdot 1,296,409.$$
All numbers that have $2$ or more factors are called \index{Numbers!composite}
{\em composite}. These include $$ 4 = 2 \cdot 2, \quad 6 = 2\cdot 3 $$ as well
as numbers that {\em look like primes}, but are in fact composite:
$$ 91 = 7 \cdot 13, \quad 161=7 \cdot 23, \quad 767 =13 \cdot 59. $$

\begin{theorem}\label{thm-pz-sqr}
Each whole number $m$ greater than $1$ possesses a lowest factor greater than
$1$. This is a prime number $p$. Unless $m$ is a prime number itself, then: $p$
is less than or equal to the square root of $m$.
\end{theorem}

All whole numbers greater than $1$ can be expressed as a product of prime
numbers --- in a unique way. This is the claim of the 1st fundamental theorem of
number theory (= fundamental theorem of arithmetic = fundamental building block
of all positive integers).

\begin{theorem}\label{thm-pz-prod}
Each element $n$ of the natural numbers greater than $1$ can be written as the
product $n = p_1 \cdot p_2 \dots p_m$ of prime numbers. If two such
factorisations $$n =  p_1 \cdot p_2 \cdots p_m = p'_1 \cdot p'_2 \cdots
p'_{m'}$$ are given, then they can be reordered such that $\;m = m'\;$ and for
all $i$:  $\;p_i = p'_i$.
\end{theorem}

In other words: each natural number other than $1$ can be written as a product
of prime numbers in precisely one way, if we ignore the order of the factors.
The factors are therefore unique (the {\em expression as a product of factors}
is unique)! For example, $$ 60 = 2 \cdot 2 \cdot 3 \cdot 5 = 2^2\cdot 3^1 \cdot
5^1. $$
And this --- other than changing the order of the factors --- is the only way in
which the number $60$ can be factorised. If you allow numbers other than primes
as factors, there are several ways of factorising integers and the uniqueness \hypertarget{uniqueness}{} is
lost: $$ 60 = 1 \cdot 60 = 2 \cdot 30 = 4 \cdot 15 = 5 \cdot 12 =6 \cdot 10 = 2
\cdot 3 \cdot 10 = 2 \cdot 5 \cdot 6 = 3 \cdot 4 \cdot 5 = \cdots . $$

The following section is aimed more at those familiar with mathematical logic:
The 1st fundamental theorem only appears to be obvious \label{remFundTheoOfArithm}. We can construct
numerous other sets of numbers (i.e. other than positive whole numbers greater
than 1), for which numbers in the set cannot be expressed uniquely as a product
of the prime numbers of the set: In the set $M = \{1, 5, 10, 15, 20, \cdots\}$
there is no equivalent to the fundamental theorem under multiplication. The
first five prime numbers of this sequence are $5, 10, 15, 20, 30$ (note: $10$ is
prime, because $5$ is not a factor of $10$ in this set --- the result is not an
element of the given basic set $M$). Because the following applies in $M$: $$
100 = 5 \cdot 20 = 10 \cdot 10 $$ and $5, 10, 20$ are all prime numbers in this
set, the expression as a product of prime factors is not unique here.

\subsection{How many prime numbers are there?}

For the natural numbers, the primes can be compared to elements in chemistry or
the elementary particles in physics (see \cite[p. 22]{Blum1999}).

Although there are only $92$ natural chemical elements, the number of prime
numbers is unlimited. Even the Greek, \index{Euclid} Euclid\footnote{Euclid,
a Greek mathematician of 4th and 3rd century before Christ. He worked at the
egyptian academy of Alexandria and wrote ''The Elements'', the most well known systematical textbook
of the Greek mathematics.} knew this in the
third century before Christ.
\begin{theorem}[Euclid]\label{thm-pz-euklid}\footnote{The common usage of the term does not denote Euclid as the inventor of the theorem rather;
the true inventor is merely not as prominent. The theorem has already been distinguished
and proven in Euclid’s Elements (Book IX, s. 20). The phraseology is remarkable due to 
the fact that the word infinite is not used. The text reads as following
$$
O\acute{\iota}~\pi\varrho\tilde{\omega}\tau o \iota~\grave{\alpha}\varrho\iota\vartheta\mu o\grave{\iota}~
\pi\lambda\varepsilon\acute{\iota}o \upsilon\varsigma~\varepsilon\grave{\iota}\sigma\grave{\iota}~
\pi\alpha\nu\tau\grave{o}\varsigma~\tau o \tilde{\upsilon}~
\pi\varrho o \tau\varepsilon\vartheta\acute{\varepsilon}\nu\tau o \varsigma~
\pi\lambda\acute{\eta}\vartheta\ o \upsilon\varsigma~
\pi\varrho\acute{\omega}\tau\omega\nu~
\grave{\alpha}\varrho\iota\vartheta\mu\tilde{\omega}\nu,
$$
the English translation of which is: the prime numbers are more than any previously existing amount
of prime numbers.}
The sequence of prime numbers does not discontinue.
Therefore, the quantity of prime numbers is infinite.
\end{theorem}
His proof that there is an infinite number of
primes is still considered to be a brilliant mathematical consideration and
conclusion today (proof by contradiction). He assumed that there is only a
finite number of primes and therefore a largest prime number. Based on this
assumption, he drew logical conclusions until he obtained an obvious
contradiction. This meant that something must be wrong. As there were no
mistakes in the chain of conclusions, it could only be the assumption that was
wrong. Therefore, there must be an infinite number of primes!

\hypertarget{euclid}{}
\paragraph{Euclid's proof by contradiction} argues as follows:

{\bf Assumption:} \quad There is a {\em finite} number of primes. \\*[4pt] {\bf
Conclusion:} \quad Then these can be listed $p_1 < p_2 < p_3 < \dots < p_n$,
where $n$ is the (finite) number of prime numbers. $p_n$ is therefore the
largest prime. Euclid now looks at the number $a = p_1 \cdot p_2 \cdots p_n +1$.
This number cannot be a prime number because it is not included in our list of
primes. It must therefore be divisible by a prime, i.e. there is a natural
number $i$ between $1$ and $n$, such that $p_i$ divides the number $a$. Of
course, $p_i$ also divides the product $a-1 = p_1 \cdot p_2 \cdots p_n$, because
$p_i$ is a factor of $a-1$. Since $ p_i $ divides the numbers $ a $ and $ a-1 $,
it also divides the difference of these numbers. Thus: $p_i$ divides  $a - (a-1)
= 1$. $p_i$ must therefore divide $1$, which is impossible. \\*[4pt] {\bf
Contradiction:} \quad Our assumption was false.

Thus there is an {\em infinite} number of primes
\hyperlink{primhfk}{(Cross-reference: overview under 2.6.1 of the number of
prime numbers in various intervals)}.\par \vskip + 3pt

Here we may mention yet another fact which is initially somewhat surprising. 
Namely, in the prime numbers sequence $p_1, p_2, \cdots,$ gaps between prime numbers can have
an individually determined length $n$. It is undeniable that under the $n$
succession of natural numbers
$$(n+1)!+2,\cdots, (n+1)!+(n+1),
$$
none of them is a prime number since in order, the numbers $2,3,\cdots,(n+1)$  
are comprised respectively as real divisors. 
($n!$ means the product of the first $n$ natural numbers therefore 
$ n!= n*(n-1)*\cdots *2*1$).

\subsection{The search for extremely large primes}

The largest prime numbers known today have several hundred
thousand digits, which is too big for us to imagine. The number of
elementary particles in the universe is ''only'' estimated to be a
$80$-digit number \hyperlink{grosord}{(See: overview under 2.6.3
of various orders of magnitude / dimensions)}.

Almost all known huge prime numbers are special candidates, called
\index{Mersenne Marin} {\em Mersenne numbers} of the form $2^p -1,$ where $p$ is
a prime. Marin Mersenne (1588-1648) was a French priest and mathematician. Not
all Mersenne numbers are prime:

$$
\begin{array}{cl}
2^2 - 1 = 3 & \Rightarrow {\rm prime} \\
2^3 - 1 = 7 & \Rightarrow {\rm prime} \\
2^5 - 1 = 31    & \Rightarrow {\rm prime} \\
   \vdots \\
2^{11} - 1 = 2.047 = 23 \cdot 89    & \Rightarrow  {\rm NOT~prime} !
\end{array}
$$

\index{Numbers!Mersenne numbers}\index{Mersenne Marin!Mersenne numbers}
\index{Mersenne Marin!theorem} Even Mersenne knew that not all Mersenne numbers
are prime (see exponent $p = 11$). A prime Mersenne number \index{Mersenne prime number} is called Mersenne prime number.  
However, he is to be thanked for the interesting conclusion that a number of the form $2^n-1$ is not a prime number
if $n$ is a composite number:

\begin{theorem}[Mersenne]\label{thm-pz-mersenne} If $2^n - 1$ is a prime number, then $n$ is also a
prime number.
\end{theorem}

\begin{Proof}{}
The theorem of Mersenne can be proved by contradiction. We therefore assume that
there exists a composite natural number $ n $ (with real factorisation)  $ n=n_1
n_2 $, with the property that $ 2^n -1 $ is a prime number.

From \begin{eqnarray*} (x^r-1)((x^r)^{s-1} + (x^r)^{s-2} + \cdots + x^r +1) & =
&  ((x^r)^s + (x^r)^{s-1} + (x^r)^{s-2} + \cdots + x^r) \\ &  & -((x^r)^{s-1} +
(x^r)^{s-2} + \cdots + x^r +1)  \\ & = & (x^r)^s -1 = x^{rs } -1,
\end{eqnarray*} we conclude \[ 2^{n_1 n_2} - 1 = (2^{n_1} -1)((2^{n_1})^{n_2 -1}
+ (2^{n_1})^{n_2 -2} + \cdots + 2^{n_1} + 1). \]
Because $ 2^n - 1 $ is a prime number, one of the above two factors on the
right-hand side must be equal to 1. This is the case if and only if $ n_1 =1 $
or $ n_2 =1$. But this contradicts our assumption. Therefore the assumption is
false. This means that there exists no composite number $ n, $ such that $ 2^n -
1 $ is a prime.
\end{Proof} \vskip + 5pt
Unfortunately this theorem only applies in one direction (the inverse does not
apply, no equivalence): see the above example $2^{11}-1, $ where $11$ is prime.

Mersenne claimed that $2^{67}-1$ is a prime number. There is also a mathematical
history behind this claim: it first took over 200 years before \index{Lucas
Edouard} Edouard Lucas (1842-1891) proved that this number is composite.
However, he argued indirectly and did not name any of the factors. Then Frank
Nelson showed in 1903 which factors make up this composite number: $$ 2^{67} -1
=147, 573, 952, 588, 676, 412, 927 = 193, 707, 721 \cdot 761, 838, 257, 287. $$
He admitted to having worked 20 years on the factorisation (expression as a
product of factors) of this 21-digit decimal number!

Due to the fact that the exponents of the Mersenne numbers do not use all
natural numbers, but only the primes, the {\em experimental space} is limited
considerably. The currently known Mersenne prime numbers \index{Mersenne prime number} 
have the exponents 
$$
\begin{array}{c}
2; ~ 3; ~ 5; ~ 7; ~ 13; ~ 17; ~ 19; ~ 31; ~ 61; ~ 89; ~ 107; ~ 127;
~ 521; ~ 607; ~ 1,279; ~ 2,203; ~ 2,281; ~ 3,217; ~ 4,253; ~ 4,423; \\
9,689; ~ 9,941, ~ 11,213; ~ 19,937; ~ 21,701; ~ 23,207; ~ 44,497; ~
86,243; ~ 110,503; ~ 132,049; ~ 216,091;\\
 756,839; ~ 859,433; ~ 1,257,787; ~ 1,398,269; ~ 2,976,221; ~ 3,021,377; ~
6,972,593; ~13,466,917.
\end{array}
$$
Thus $39$ Mersenne prime numbers are currently known.

The $19$th number with the exponent $4,253$ was the first with at least $1,000$ digits in  decimal system
(the mathematician Samual \index{Yates Samual} Yates coined the expression {\em
titanic} \index{Prime number!titanic} prime for this; it was discovered by
Hurwitz in 1961); the $27$th number with the exponent $44,497$ was the first
with at least $10,000$ digits in the decimal system (Yates coined the expression
\index{Prime number!gigantic}  {\em gigantic} prime for this. These names are
now long outdated).

The 37th number was found in January 1998 and has 909,526
digits in the decimal system, which corresponds to 33 pages in the newspaper!

\newpage
These numbers can be found at the following URLs:

{\href{http://reality.sgi.com/chongo/prime/prime_press.html}{\tt http://reality.sgi.com/chongo/prime/prime\_press.html}}
\vskip - 4pt \noindent
\begin{quote}
(The supercomputer manufacturer SGI Cray Research not only employed brilliant
mathematicians but also used the prime number tests as benchmarks for their
machines.)
\end{quote}
{\href{http://www.utm.edu/}{\tt http://www.utm.edu/}}
\vskip - 4pt \noindent
\begin{quote}
(At the university of Tennessee you will find extensive research results about
prime numbers.)
\end{quote}
\subsubsection{M-38 -- June 1999}\index{Mersenne Marin!M-38}
The 38th Mersenne prime, called M-38,
$$ 2^{6,972,593} - 1. $$
was discovered in June 1999 and has $2,098,960$ digits in the decimal system
(that corresponds to around 77 pages in the newspaper).

\subsubsection{M-39 -- December 2001} \index{Mersenne Marin!M-39}

The 39th Mersenne prime, called M-39, $$2^{13,466,917}-1,$$ was found at December 6, 2001 –- more exactly,
the verification of this number, found at November 14, 2001 by the Canadian student Michael Cameron,
was successfully finished. This number has about 4 million decimal digits (exactly 4,053,946 digits).
Trying only to print this number $(924947738006701322247758 \cdots 1130073855470256259071)$ in the Financial Times
would need around 200 pages.

\subsubsection{GIMPS}\index{GIMPS}

Discovering the 39th Mersenne prime the GIMPS project (Great Internet Mersenne Prime Search) founded in 1996
already discovered for the 5th time the greatest Mersenne number to be proven being prime.

Now more than 130,000 volunteers, amateurs and experts, are working for the GIMPS project. They connect their
computers into the so called ''primenet'', organized by the company entropia, to find such numbers using distributed
computer programs.

\subsubsection{EFF}\index{EFF}

This search is also spurred on by a competition started by the non-profit
organisation EFF (Electronic Frontier Foundation) using the means of an unknown
donator. The participants are rewarded with a total of 500,000 USD if they find
the longest prime number. In promoting this project, the unknown donator is not
looking for the quickest computer, but rather wants to draw people's attention
to the possibilities offered by {\em cooperative networking}

{\href{http://www.eff.org/coopawards/prime-release1.html}{\tt http://www.eff.org/coopawards/prime-release1.html}}.

The discoverer of M-38 got for discovering the first prime with more than 1 million decimal digits 50,000 USD
from the EFF. The next price from EFF with 100,000 USD is for a proven prime with more than 10 million decimal digits.
%{\href{http://www.octocad.demon.co.uk/mersenne/prime.htm }{\tt http://www.octocad.demon.co.uk/mersenne/prime.htm}}.

Edouard Lucas (1842-1891) held the record for the longest prime number for over
70 years by proving that $2^{127}-1$ is prime. No new record is likely to last
that long.

\subsubsection{Prime number tests}
\index{Prime number!test}
In order to implement secure encryption procedures we need extremely large prime
numbers (in the region of $2^{2,048}$, i.e. numbers with $600$ digits in the
decimal system!).

Up to now we have always looked for the prime factors in order to decide whether
a number is prime. However, if even the smallest prime factor is enormous, the
search takes too long. Factorising numbers using systematic computational
division or using the sieve of Eratosthenes (see below) is only feasible using
current computers for numbers with up to around $20$ digits in the decimal
system.

However, if we know something about the {\em construction} of the number in
question, there are extremely highly developed procedures that are much quicker.
In the 17th century, Fermat wrote to Mersenne that he presumed that all numbers
of the form $$ F(n) = 2^{2^n} + 1 $$ are prime for all whole numbers $n$ greater
than or equal to $0$ (see below).

As early as in the 19th century, it was discovered that the $29$-digit number $$
F(7) = 2^{2^7} + 1 $$ is not prime. However, it was not until 1970 that
Morrison/Billhart managed to factorise it.
\begin{eqnarray*}
F(7) & = & 34,279,974,696,877,740,253,374,607,431,768,211,457 \\
& = & 59, 649, 589, 127, 497, 217 \cdot  574, 689, 200, 685,
129,054, 721.
\end{eqnarray*}
Many rapid prime number tests are based on the (small) Fermat theorem put
forward by Fermat in 1640 .
\index{Fermat!small Fermat}
\begin{theorem}[''small'' Fermat]\label{thm-pz-fermat1}
Let $p$ be a prime number and $a$ be any whole number, then for all $a$ $$a^p
\equiv a \; {\rm mod} \; p.$$
This could also be formulated as follows: \\ Let $p$ be a prime number and $a$
be any whole number that is not a multiple of $p$ (also $a \not\equiv 0 \; {\rm
mod} \; p$), then $a^{p-1} \equiv 1 \; {\rm mod} \; p$.
\end{theorem}

If you are not used to calculating with remainders (modulo), please simply
accept the theorem. What is important here is that this sentence implies that if
this equation is not met for any whole number $a$, then $p$ is not a prime! The
tests (e.g. for the first formulation) can easily be performed using the {\em
test basis} $a = 2$.

This gives us a criterion for non-prime numbers, i.e. a negative test, but no
proof that a number $a$ is prime. Unfortunately Fermat's theorem does not apply
--- otherwise we would have a simple proof of the prime number property (or to
put it in other words, we would have a simple prime number criterion).

Comment: Numbers that have the property $$ 2^n \equiv 2 \;{\rm mod}\; n $$ but
are not prime are called \index{Pseudoprime number}\index{Numbers!pseudoprime
number} {\em pseudoprime numbers}. The first pseudoprime number (i.e. not a
prime) is $$ 341 = 11 \cdot 31. $$
There are numbers that pass the Fermat test with all bases and yet are not
prime: these numbers are called \index{Numbers!Carmichael numbers} {\em
Carmichael numbers}. The first of these is $$ 561 = 3 \cdot 11 \cdot 17. $$
A stronger test is provided by \index{Miller} \index{Rabin} Miller/Rabin: it is
only passed by so-called {\em strong pseudoprime numbers}. Again, there are
strong pseudoprime numbers that are not primes, but this is much less often the
case than for (simple) pseudoprime numbers. The smallest strong pseudoprime
number base $2$ is $$ 15,841 = 7 \cdot 31 \cdot 73. $$
If you test all 4 bases, $2, 3, 5$ and $7$, you will find only one strong
pseudoprime number up to $25 \cdot 10^9$, i.e. a number that passes the test and
yet is not a prime number.

More extensive mathematics behind the Rabin test delivers the probability that
the number examined is prime (such probabilities are currently around $10^{-
60}$).

Detailed descriptions of tests for finding our whether a number is prime can be
found on Web sites such as:

{\href{http://www.utm.edu/research/primes/mersenne.shtml}{\tt
http://www.utm.edu/research/primes/mersenne.shtml}} and

{\href{http://www.utm.edu/research/primes/prove/index.html}{\tt
http://www.utm.edu/research/primes/prove/index.html}}.

\subsection{The search for a formula for prime numbers}
\index{Prime number!formula}
There are currently no useful, open (i.e. not recursive) formulae known that
only deliver prime numbers (recursive means that in order to calculate the
function the same function is used with a smaller variable). Mathematicians
would be happy if they could find a formula that leaves gaps (i.e. does not
deliver all prime numbers) but does not deliver any composite (non-prime)
numbers.

Ideally, we would like, for the number $n$, to immediately obtain the $n$-th
prime number, i.e. for $f(8) = 19\,$ or for  $f(52) = 239$.

Ideas for this can be found at

{\href{http://www.utm.edu/research/primes/notes/faq/p_n.html}
 {\tt http://www.utm.edu/research/primes/notes/faq/p\_n.html}}.

Cross-reference:  \hyperlink{ntePrimzahl}{the table under 2.6.2} contains the precise values for the $n$th
prime numbers for selected $ n.$

The following are some of the most common ideas for prime number formulae:
\begin{enumerate}
  \item Mersenne numbers  $f(n) = 2^n - 1$: \\
    As shown above, this formula seems to deliver relatively large prime numbers
but - as for $n=11$ [$f(n)=2,047$] - it is repeatedly the case that the result
is not prime.
  \item $F(k,n) = k \cdot 2^n \pm 1$  for $ n $ prime and $ k $ small primes:\\
   For this generalisation of the Mersenne numbers there are (for small $k$)
also extremely quick prime number tests (see \cite{Knuth1981}). This can be
performed in practice using software such as the PROTHS software from Yves
Gallot

({\href{http://www.prothsearch.net/index.html}{\tt
http://www.prothsearch.net/index.html}}).

  \item {\em Fermat numbers}\footnote{The Fermat prime numbers play a role in circel division.
As proven by Gauss\index{Gauss}, a regular $p$-edge
can only be constructed with the use of a pair of compasses and a ruler, when $p$ is a Fermat prime number.}
\index{Numbers!Fermat numbers}\index{Fermat!Fermat
number}  $F(n) = 2^{2^n} + 1$:\\
As mentioned above, Fermat wrote to Mersenne regarding this assumption.
Surprisingly he would have been able obtain a positive result using the negative
prime number test for $n=5$ based on his small theorem.
$$
\begin{array}{lll}
F(0) = 2^{2^0} + 1  = 2^1 + 1 & = 3 &   \mapsto {\rm ~prime}  \\
F(1) = 2^{2^1} + 1  = 2^2 + 1 & = 5 &   \mapsto {\rm ~prime}  \\
F(2) = 2^{2^2} + 1  = 2^4 + 1 & = 17 &  \mapsto {\rm ~prime}  \\
F(3) = 2^{2^3} + 1  = 2^8 + 1 & = 257 & \mapsto {\rm ~prime}  \\
F(4) = 2^{2^4} + 1  = 2^{16} + 1 &  = 65,537 &  \mapsto {\rm ~prime}  \\
F(5) = 2^{2^5} + 1  = 2^{32} + 1 &  = 4 ,294, 967, 297 = 641 \cdot
6,700,417 &  \mapsto {\rm ~NOT~prime} !
\end{array} 
$$
  \item Carmichael numbers: see above.
  \item Pseudoprime numbers: see above.
  \item Strong pseudoprime numbers: see above.
  \item Idea based on \hyperlink{euclid}{Euclid's proof} (infinite many prime
numbers) $p_1 \cdot p_2 \cdots p_n +1$:

$$
\begin{array}{lll}
2{\cdot}3 +1 &      = 7 &          \mapsto {\rm ~prime} \\
2{\cdot}3{\cdot}5 +1 &      = 31    &      \mapsto {\rm ~prime} \\
2{\cdot}3{\cdot}5{\cdot}7 +1 &      = 211   &      \mapsto {\rm ~prime} \\
2{\cdot}3{\cdots}11 +1 &        = 2,311  &      \mapsto {\rm ~prime} \\
2\cdot3 \cdots 13 +1 &  = 59 \cdot 509 &    \mapsto {\rm ~NOT~prime} ! \\
2\cdot3 \cdots 17 +1 &  = 19 \cdot 97 \cdot 277 &   \mapsto {\rm ~NOT~prime} !
\\
\end{array}
$$

  \item As above but except $+1$: $p_1 \cdot p_2 \cdots p_n -1$

$$
\begin{array}{lll}
2\cdot 3 -1     &   = 5 &   \mapsto {\rm ~prime} \\
2\cdot 3 \cdot  5  -1   &   = 29 &  \mapsto {\rm ~prime} \\
2\cdot 3 \cdots 7  -1   &   = 11 \cdot 19 & \mapsto {\rm ~NOT~prime} ! \\
2\cdot 3 \cdots 11 -1   &   = 2,309 &    \mapsto {\rm ~prime} \\
2\cdot 3 \cdots 13 -1   &   = 30,029 &    \mapsto {\rm ~prime} \\
2\cdot 3 \cdots 17 -1    &  = 61 \cdot 8,369 &   \mapsto {\rm ~NOT~prime!}
\end{array} 
$$

  \item \index{Euclidean number} {\em Euclidean numbers} $e_n = e_0 \cdot e_1
\cdots e_{n-1} + 1$  with $n$ greater than or equal to $1$ and $e_0 := 1$.
$e_{n-1}$ is not the $(n-1)$th prime number, but the number previously found
here.
    Unfortunately this formula is not open but recursive.
    The sequence starts with

>>>>

@@

One encryption procedure has been proved to be secure --- the {\em One Time
  Pad}.
\index{One ­Time ­Pad} However, this procedure has several practical
disadvantages (the key used must be selected randomly and must be just as long
as the message to be protected), which means that it is hardly used except in
closed environments such as for the hot wire between Moscow and Washington.\par \vskip + 3pt

@@

The most well-known symmetric encryption procedure is the \index{DES} DES-algorithm. The DES-algorithm has been developed by IBM in collaboration with the
National Security Agency \index{NSA} (NSA), and was published as a standard in
1975. Despite the fact that the procedure is relatively old, no effective attack
on it has yet been detected. The most effective way of attacking consists of
testing all possible keys until the right one is found ({\em brute--force--attack}).
\index{Attack!brute-force} Due to the relatively short key length of
effectively 56 bits (64 bits, which however include 8 parity bits), numerous
messages encrypted using DES have in the past been broken. Therefore, the
procedure can now only be considered to be conditionally secure. Symmetric
alternatives to the DES procedure include the IDEA \index{IDEA} or Triple DES
algorithms.\par \vskip + 3pt

@@
